Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-09-07 ms.tex
analysis figures updated, text to follow

Moreover, a substantial value of $X_\CR$ can bias hydrostatic cluster masses as well as the temperature decrement of the CMB in the direction of a galaxy cluster due to the Sunyaev-Zel'dovich effect which is proportional to the line-of-sight integral of the thermal electron pressure. This could then severely jeopardize the use of clusters to determine cosmological parameters. Comparing X-ray and optical potential profiles in the centers of galaxy clusters yields an upper limit of 20--30 \% of non-thermal pressure (that can be composed of CRs, magnetic fields or turbulence) relative to the thermal gas pressure \citep{article:Churazov_etal:2008, article:Churazov_etal:2010}. An analysis that compares spatially resolved weak gravitational lensing and hydrostatic X-ray masses for a sample of 18 galaxy clusters detects a deficit of the hydrostatic in comparison to the lensing mass of $20 \%$ at $R_{500}$ -- within which the
mean density is 500 times the critical density -- suggesting again a substantial non-thermal pressure contribution on large scales \citep{article:Mahdavi_etal:2008}. High-energy gamma-ray emission is a complementary method of constraining the pressure contribution of CRs that is most sensitive
to the cluster core region. However, it assumes that the CR component is fully mixed with the ICM and almost insensitive in detecting a two-phase structure of CRs and the thermal ICM.  A fraction $X_\CR$ of only a few percent is required in order to produce a gamma-ray flux observable with current generation
gamma-ray telescopes rendering this technique at least as sensitive as the dynamical and hydrostatic method (which are however more general in that they are sensitive to any non-thermal pressure component).

Gamma-ray emission can also be produced by Compton up-scattering of ambient photons, for example cosmic microwave background (CMB) photons, on ultra-relativistic leptons. Those leptons can either be secondaries from the above mentioned CR interactions or be injected into the ICM by powerful cluster
members and be further accelerated by diffusive shock acceleration or turbulent reacceleration processes.

@@

@@

\section{Constraining the Magnetic Field}
\label{sec:B}
We have seen in the previous section that we can obtain an absolute lower limit on the gamma-ray emission in the hadronic model by assuming high magnetic fields, $B>B_\rmn{CMB}$. Hence, we can turn this around by using our upper limit on the gamma-ray emission (and by extension on the CR pressure) to infer a lower limit on the magnetic field that is needed to explain the observed radio emission which, however, assuming the validity of the hadronic model \citep{article:PfrommerEnsslin:2004a}. Lowering the gamma-ray constraint will tighten (increase) the magnetic field limit. Once this gets in conflict with magnetic field measurements by means of other methods, e.g., Faraday rotation measure (RM)\footnote{Generally, Faraday RM analysis of the magnetic-field strength by, e.g., background sources through clusters are degenerate with the magnetic coherence scale and may be biased by the unknown correlation between magnetic and density fluctuations.}, this would challenge the hadronic model of radio halos. The method that we use to constrain the magnetic field inherits a dependence on the assumed spatial structure that we parametrize by
\begin{equation}
\label{eq:B}
B(r) = B_{0} \,\left(\frac{n_{\rmn{e}(r)}}{n_{\rmn{e}(0)}}\right)^{\alpha_B},
\end{equation}
as suggested by Faraday RM studies and numerical magneto-hydrodynamical (MHD) simulations \citep[][and references therein]{article:Bonafede_etal:2010, article:Bonafede_etal:2011}. In fact, the magnetic field in the Coma cluster is among the best constrained, because its proximity permits RM observations of 
seven radio-sources located at projected distances of 50 to 1500 kpc from the cluster center. The best-fit model obtains $B_{0} = 4.7^{+0.7}_{-0.8}\,\mu$G and $\alpha_{B} = 0.5^{+0.2}_{-0.1}$ \citep{article:Bonafede_etal:2010}. We aim for constraining the central field strength, $B_{0}$, which depends on the magnetic decline, $\alpha_{B}$, that we vary (within reasonable ranges of $\Delta\alpha_{B}=0.2$ as suggested from the Faraday RM studies) to address this dependence. We proceed as follows:

@@

\begin{figure*}
\begin{center}
%\scalebox{1.0}{\plotone{f4.eps}}
\end{center}
\caption{Integral flux upper limits (this work, Table X) compared with simulated integrated spectra of VHE gamma-ray emission from the Coma cluster.}
\label{fig:spectrum}
\end{figure*}

>>>>

Moreover, a substantial value of $X_\CR$ can bias hydrostatic cluster masses as well as the temperature decrement of the cosmic microwave background (CMB) in the direction of a galaxy cluster due to the Sunyaev-Zel'dovich effect which is proportional to the line-of-sight integral of the thermal electron pressure. This could then severely jeopardize the use of clusters to determine cosmological parameters. Comparing X-ray and optical potential profiles in the centers of galaxy clusters yields an upper limit of 20--30 \% of non-thermal pressure (that can be composed of CRs, magnetic fields or turbulence) relative to the thermal gas pressure \citep{article:Churazov_etal:2008, article:Churazov_etal:2010}. An analysis that compares spatially resolved weak gravitational lensing and hydrostatic X-ray masses for a sample of 18 galaxy clusters detects a deficit of the hydrostatic in comparison to the lensing mass of $20 \%$ at $R_{500}$ -- within which the
mean density is 500 times the critical density -- suggesting again a substantial non-thermal pressure contribution on large scales \citep{article:Mahdavi_etal:2008}. High-energy gamma-ray emission is a complementary method of constraining the pressure contribution of CRs that is most sensitive
to the cluster core region. However, it assumes that the CR component is fully mixed with the ICM and almost insensitive in detecting a two-phase structure of CRs and the thermal ICM.  A fraction $X_\CR$ of only a few percent is required in order to produce a gamma-ray flux observable with current generation
gamma-ray telescopes rendering this technique at least as sensitive as the dynamical and hydrostatic method (which are however more general in that they are sensitive to any non-thermal pressure component).

Gamma-ray emission can also be produced by Compton up-scattering of ambient photons, for example CMB photons, on ultra-relativistic leptons. Those leptons can either be secondaries from the above mentioned CR interactions or be injected into the ICM by powerful cluster
members and be further accelerated by diffusive shock acceleration or turbulent reacceleration processes.

@@

%This is more constraining by 50\% in comparison to the model derived from simulations. (in comparison to a power law spectrum with $\alpha_\rmn{TeV} \sim 2.2$ down to a cutoff in the normalized momentum of 0.8)

@@

\section{Constraining the Magnetic Field}
\label{sec:B}
We have seen in the previous section that we can obtain an absolute lower limit on the gamma-ray emission in the hadronic model by assuming high magnetic fields, $B>B_\rmn{CMB}$. Hence, we can turn this around by using our upper limit on the gamma-ray emission (and by extension on the CR pressure) to infer a lower limit on the magnetic field that is needed to explain the observed radio emission which, however, assuming the validity of the hadronic model \citep{article:PfrommerEnsslin:2004a}. Lowering the gamma-ray constraint will tighten (increase) the magnetic field limit. Once this gets in conflict with magnetic field measurements by means of other methods, e.g., Faraday rotation measure (RM)\footnote{Generally, Faraday RM analysis of the magnetic-field strength by, e.g., background sources through clusters are degenerate with the magnetic coherence scale and may be biased by the unknown correlation between magnetic and density fluctuations.}, this would challenge the hadronic model of radio halos. The method that we use to constrain the magnetic field inherits a dependence on the assumed spatial structure that we parametrize by
\begin{equation}
\label{eq:B}
B(r) = B_{0} \,\left(\frac{n_{\rmn{e}}(r)}{n_{\rmn{e}}(0)}\right)^{\alpha_B},
\end{equation}
as suggested by Faraday RM studies and numerical magneto-hydrodynamical (MHD) simulations \citep[][and references therein]{article:Bonafede_etal:2010, article:Bonafede_etal:2011}. In fact, the magnetic field in the Coma cluster is among the best constrained, because its proximity permits RM observations of 
seven radio-sources located at projected distances of 50 to 1500 kpc from the cluster center. The best-fit model obtains $B_{0} = 4.7^{+0.7}_{-0.8}\,\mu$G and $\alpha_{B} = 0.5^{+0.2}_{-0.1}$ \citep{article:Bonafede_etal:2010}. We aim for constraining the central field strength, $B_{0}$, which depends on the magnetic decline, $\alpha_{B}$, that we vary (within reasonable ranges of $\Delta\alpha_{B}=0.2$ as suggested from the Faraday RM studies) to address this dependence. We proceed as follows:

@@

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f4.eps}}
\end{center}
\caption{Integral flux upper limits (this work, Table X) compared with simulated integrated spectra of VHE gamma-ray emission from the Coma cluster.}
\label{fig:spectrum}
\end{figure*}
